---
title: "README"
author: "Matthew Zinkgraf"
date: "10/31/2018"
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
#library(BiometricsWWU, quietly = TRUE)
```

# BioStatsWWU

BioStatsWWU is a R package meant to help support Biostatistics classes (BIOL340 and ESCI340) at Western Washington University (WWU) and is available on [GitHub](https://github.com/mzinkgraf/BioStatsWWU).

The BioStatsWWU Package can be installed from github:

```{r, eval = F}
    #install dependencies
    install.packages(c("devtools","prob","plyr","reshape2","ggplot2"),
                     dependencies = TRUE)

    #install BiometricsWWU
    devtools::install_github("mzinkgraf/BioStatsWWU")
```

---

# Motivation

[todo]


# Further Reading

## Online Resources

- <https://whitlockschluter3e.zoology.ubc.ca/>: Website of "The analysis of biological data" by Michael Whitlock and Dolph Schluter.
- [WWU Biometrics Website](https://mzinkgraf.github.io/BioStatsWWU/): This site is for students taking BIOL 340 or ESCI 340 Biostatistics class at Western Washington University.


