---
title: "Data Types"
author: "Ben Miner and Matthew Zinkgraf"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Data Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteRender{all}
---

[Back to Weekly Labs](Weekly_labs.html)

<h1>Data in R</h1>

# Data Structure

An important part of science is entering and managing data. Because we
do most of our statistics with computers, we have to enter and store
data on computers. Although this sounds simple, many students have no
idea how to properly store data.

Generally a spreadsheet program is used to enter and store data. The
important thing to remember is start in the upper left-hand corner of
the spreadsheet and use the first row to label each column. Because some
programs do not like spaces, I always use an underscore "\_" or "."
instead of a space. For example, I might label a column that has the
initial weight in grams as "Initial_weight_g". It is important to
provide a column for each piece of important information. For example,
if you manipulated temperature and light in your experiment you should
have a column for temperature and a column for light--not a column for
treatment, though you might include a third column label treatment.

Once you have labeled the columns you can begin to enter data. Make sure
to always enter something for each column, unless there is a reason to
leave a cell blank. I will provide an example spreadsheet to illustrate
how to properly enter data into a spreadsheet.

# Data in R

Because we need to get data into R for analyses, it is important to know
the objects that can store data in R: vectors, , factors, data.frames,
arrays, matrices, and lists. We have already seen examples of vectors.
For example, <samp>1:10</samp> produces a vector with 10 integers
beginning with 1 and ending with 10. Let's go over each of these
objects, but keep in mind that vectors, factors, and data.frames are
used most often for storing data.

## Common data types for graphing and analyzing

### Vector

Vectors are a set of values that all have the same type. For example,
all the values in a vector might be integers. Vectors cannot be a
combination of different types of values (e.g., a group of words and
numbers), and R will either give you an error, or convert all the values
to a common type. You will use vectors a lot because they often
represent your data in R. So it is important that we know how to create
them.

There are several ways to create vectors in R.

<dl class="dl-horizontal">

<dt>

c

</dt>

<dd>

This function concatenates numbers or characters separated by commas.

</dd>

<dt>

sequence

</dt>

<dd>

This function generates a vector of integers starting at 1 and counting
by ones.

</dd>

<dt>

seq

</dt>

<dd>

This function creates a sequence of numbers, but is more flexible than
the function sequence. Putting a colon between two numbers is short-hand
for this function when you are counting by ones.

</dd>

</dl>

Try and generate a few vectors with these functions. Use the help file
to see which arguments are required for each function. Also use the
colon to quickly generate a vector of integers. Below are a few examples
without the output, so you can think about what the output should look
like before you run each example.

```{r, eval = FALSE}
c(1, 5, 10.3)
sequence(100)
4:12
-34:28
seq(4, 12)
seq(-34, 28)
seq(4, 12, by = 0.2)
seq(1, 200, length.out = 5)
seq(1, 100, along.with = 1:10)
c("Ben", "Melissa", "Gabriel", "Gavin") #a vector of characters
```

Here are a few more example with the output. Remember that all values of
the vector must have the same type (e.g., numbers).

```{r}
c(1, 2, 3.4, 6, 8)
2:6
seq(from = 4, to = 5, by = 0.1)
rep(c(4, 5), 3)
rep(c(4, 5), each = 3)
```

The code below illustrates how R will convert all items in a vector to
the same data type. Notice that I include both numbers and characters in
the arguments for function <samp>c()</samp>. Now look at the output and
notice that all items have quotes around them, which indicates that R
has converted the numbers to characters. In other words,
<samp>"4"</samp> is not the number 4, but the character four.

```{r}
c(4, 3, "Ben", "Dexter", "orange")
```

### Factor

A factor is a vector of data in which there are specific groups in the
vector. It is typically used to tell R that the variable is categorical.
For example, an experiment where we have two treatments, light and dark,
and these treatments represent a categorical variable. Making sure R, or
any other statistical program, knows which variables are categorical is
very important so the computer uses the correct test. The function
<samp>factor()</samp> is a common way to create a factor.

```{r}
#Create a vector
(treatments <- rep(c("light", "dark"), each = 4))
#Convert the vector to a factor
(treatment.fac <- factor(treatments)) #Assigns and prints vector
#create another vector
vals <- rep(LETTERS[1:2], each = 5); vals
#Convert the vector to a factor
ben.factor <- factor(vals); ben.factor #Also assigns and prints vector
```

There is also another function <samp>gl()</samp> that you can use to
quickly create a factor. This function requires you indicate the number
of levels <samp>n</samp>, the number of replicates in each level
<samp>k</samp>, and the names of each level <samp>labels</samp>. This
function is more concise, and creates a vector and converts it to a
factor at the same time.

```{r}
#More concise method with gl()
(treatment.fac2 <- gl(n = 2, k = 4, labels = c("light", "dark"))) #Same as treatment.fac
(new.factor <- gl(3, 10, labels = c("low", "med", "high")))
```

By default R doesn't print the object when you assign it a name. To
print the object, just type the name of the object or put paratheses
around code that assigns a name. Both are illustrated above. The
semicolon is use when you want to put additional code on a single line.

What do you notice that is different between the variable
<samp>vals</samp> and the variable <samp>ben.factor</samp>? There are
lots of advantages to using factors. For example, it is easy to perform
calculations on groups specified by the factor. For example, say the
<samp>ben.factor</samp> specifies two treatments in an experiment, and
our data are as follows.

```{r}
#Enter in some numeric data to go with ben.factor
data <- c(1, 2, 3, 2, 3, 5, 4, 6, 7, 6) 
length(data)    #Just checking sample size of data
```

It is now very easy to calculate, say the mean and standard deviation
for treatment "A" and treatment "B". To do this we just use the
<samp>tapply()</samp> function. In the parentheses, first give the
vector with the numbers (in this case <samp>data</samp>), next give the
factor (in this case <samp>ben.factor</samp>), and last give the
function you want to apply (in this case <samp>mean</samp>). Try this

```{r}
tapply(data, ben.factor, mean)
tapply(data, ben.factor, sd)
```

### Data.frame

<samp>data.frames</samp> are really just a collection of vectors and
factors, and represent exactly what you see in an Excel spreadsheet. In
other words, they represent a table of data with column labels at the
top. <samp>data.frames</samp> are easy to make with the function
<samp>data.frame()</samp>, where the different vectors or factors,
separated with commas, go in the parentheses. For example, let's make a
data.frame with our variables we have already created.

```{r}
ben.frame <- data.frame(Treatment = ben.factor, Data = data)
ben.frame
```

We can check to see what type of variable each column represents with
the function <samp>str()</samp>, which stands for structure.

```{r}
str(ben.frame)
```

We can also get a summary of our data with the function
<samp>summary()</samp>.

```{r}
summary(ben.frame)
```

Now create your own <samp>data.frame</samp> with two factors, and one
numeric vector.

## Other useful data types

### Matrix

The function <samp>matrix()</samp> is the primary way to create a
matrix. Like a vector, all members of a matrix must be the same type.
There are three arguments, <samp> nrow</samp>, <samp>ncol</samp>, and
<samp>byrow</samp>, that are commonly used when creating a matrix with
the function <samp>matrix()</samp>. You can use either <samp>nrow</samp>
to indicate the number of rows (R then calculates the number of
columns), or <samp>ncol</samp> to indicdate the number of columns (R
then calculates the number of rows). The argument <samp>byrow</samp>, by
default is <samp>FALSE</samp>, indicates whether you want to fill in
rows or columns first. If <samp>byrow = FALSE</samp>, R will fill in all
the rows of column 1 before filling in values in the column. If
<samp>byrow = TRUE</samp>, R will fill in all the values of row 1 before
filling in values in the second row.

```{r}
matrix(1:10, nrow = 2)
matrix(1:10, ncol = 2)
matrix(1:10, nrow = 2, byrow = T)
```

### Array

An array is similar to a matrix, but you can specify some addtional
information, like the dimensions. The function <samp>array()</samp> is
the primary way to create an array. An array is different from a matrix
because a matrix allows specific mathematical operations that you cannot
do on an array. Typically programmers use an array to store information
as a program runs. For example, you create a loop to do a calculation
many times, and the result of each calculation is stored in an array.

```{r}
array(3:12, dim = c(2, 5))

#example of looping and using an array.
my.array <- array(NA, dim = c(2, 10))
my.array
for(i in 1:2){
  for(j in 1:10){
    my.array[i, j] <- i*j
  }
}
my.array
```

### List

A list allows you to group different types of data into a single list.
For example, a list can have characters, numbers, dates, vectors, etc.
The function <samp>list()</samp> is the primary way to create a list.
You can also name part of a list.

```{r}
list("Ben", 1:5, letters[1:4])
#Now with names
list(Name = "Ben", Numbers = 1:5, Letters = letters[1:4]) 
```

<samp>letters</samp> is a built-in vector of the lower-case alphabet,
and <samp>LETTERS</samp> is a vector of the upper-case alphabet.
