---
title: "plyr"
author: "Ben Miner and Matthew Zinkgraf"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{plyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteRender{all}
---

<form>
 <input type="button" value="Back" onclick="history.back()">
</form>

<h1>Plyr Package</h1>

The plyr package for R helps summarize data quickly.  For example, have you ever tried to calculate the means for a bunch of different groups in Excel; what a pain in the ass.  We have already learned the <samp>tapply()</samp> function, but it limits you to only one summary stat (e.g., the means for each group of interest).  The plyr package shines when you want to summarize several different stats.

# Install and Load plyr

Use the <samp>install.package()</samp> to install the package.  Then use either the <samp>library()</samp> or <samp>requires()</samp> functions to load the package.  Alternatively, in RStudio click on the Packages tab in the lower-right window, and then the Install button on the left-hand side just below the tabs.  Once the package is installed, just check the box next to the plyr package.  Remember you need to only install the package on a computer once, but need to load the package each time you restart R.  Look at the page on packages for more information.  

```{r, eval=FALSE}
  #If you have admin access
  install.packages("plyr", dependencies=T)
  require("plyr")

  #If you don't have admin access
  #And install the package to your u drive
  install.packages("plyr", lib="u:/", dependencies=T)
  require("plyr", lib.loc="u:/")
```

# ddply

There many functions in this package, but I want to focus on the function <samp>ddply()</samp>.  This function take a data.frame, summarizes it, and returns a data.frame to the user.  There are other functions to summarize other data types, and you can even convert between data types. For example, the function <samp>laply()</samp> take a list, summarizes it, and returns an array.  

The function <samp>ddply()</samp> requires several arguments.  The first is the data.frame that you want to summarize.  The second is the columns that you want to summarize by.  There is a bunch of data that R already installed on your computer, and we are going to just look at the <samp>iris</samp> data in R to see how this function works.  If you want to see all the data available in R, use the function <samp>data()</samp> without any arguments.  Here I load the the cars data and look at the structure of it.  

```{r, echo=FALSE, message=FALSE, warning=FALSE}
require(plyr)
```

```{r}
  data(iris)
  str(iris)
```

So, let us first just ask what was the average and standard deviation of rbi per team.  

```{r}
sepal.length.species <- ddply(iris, .(Species), summarise, 
  mean.Sepal.Length = mean(Sepal.Length, na.rm = T),
  sd.Sepal.Length = sd(Sepal.Length, na.rm = T)
)
sepal.length.species
```

I make up the names <samp>mean.Sepal.Length</samp> and <samp>sd.Sepal.Length</samp>.  If I wanted to caculate the variance, then I could add another argument of change one of the last two arguments.  Here I will just add another.  

```{r}
  sepal.length.species2 <- ddply(iris, .(Species), summarise, 
    mean.Sepal.Length = mean(Sepal.Length, na.rm = T),
    sd.Sepal.Length = sd(Sepal.Length, na.rm = T),
    var.Sepal.Length = var(Sepal.Length, na.rm = T)
  )
sepal.length.species2
```

If there were more categorical variables in this data.frame, then we could summarize by them also.  To show you what I mean, let's add another factor to the iris data.frame.

```{r}
iris2 <- data.frame(iris, NewFactor = rep(c("Big", "Small"), length.out=iris))
head(iris2)
```

Now let's summarize by species and the new factor we just created.  

```{r}
  sepal.length.species.newfac <- ddply(iris2, .(Species, NewFactor), summarise, 
    mean.Sepal.Length = mean(Sepal.Length, na.rm = T),
    sd.Sepal.Length = sd(Sepal.Length, na.rm = T)
  )
sepal.length.species.newfac
```
